Towards a wide-coverage grammar for Swedish

### 2011-12-19 doc/report/report.tex
report: grammar section revised

\chapter{The grammar}
\label{sec:prog.grammar}
An important part of this project has been to developed a grammar and
adapting it to cover
constructions used in Talbanken. A grammar can never be expected to give
full coverage of a language, but we aim for a grammar fragment which gives a deep
analysis of the most important Swedish constructions.
The starting point has been the GF resource grammar and the new implementation is
still compatible with this.
Before describing the work progress we will give an introduction to the
resource grammars in general and to the Swedish implementation in particular.
This is found in section \ref{sec:swegf} while section \ref{sec:Added} shows and
explains the development during this project.

@@

\section{The Swedish resource grammar}
%>> start by easy examples, jag ser inte dig, du ser inte mig osv. 
%>> plocka från svenska delen 1 2 4 
%>> show simple trees
%>> explain all in Sw. section.
%>> show abstract and concrete tree, compare. (or in Gf sect)
%>> no subjects or object, analyse is more from the lexicon,
%>> compare to others, ask elisabet.\\
\label{sec:swegf}
%>> gf has pos-driven, leads to part-of-sentence analyse. first have v2,a2,vs etc.
%>> they may be combined in traditional way, and so the types depend and relate to each other
%>> see slides from summer school.
>> functions for predication, complementation.
>> rather different from others syntactical theories. does not remember syntactical
>> parts of sentences, like subj, obj, attributes.
%>> worked well until now, but particular langs. needs particular solutions. --> reflexives!
%>> no syntactic theory can cover a whole language, and so gf can't either.
>> shares 80\% with scandinavian, mostly words that differ.\\
>> complete morphology, limited syntax
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other Scandinavian languages.
Apart form describing morphology, the resource grammar deals with
word order, agreement, tense, basic conjunction etc.

The analysis preformed by GF is driven by the parts of speech, which are combined into
parts of sentences.
Figure \ref{fig:gfpic} shows the different categories, or types, used in the
resource grammars. 
A lexicon normally consist of words from the open word-classes, shown in rectangular boxes in
the picture. Each word has a type describing its word class, and the abstract grammar
consists of rules for how the parts of speech can be used -- how the types relate
to each other.
\begin{figure}[h]
\includegraphics[height=90mm]{categories.png}
\caption{The types of the GF resource grammars.}
\label{fig:gfpic}
\end{figure}

A category is represented by a table covering the forms the phrase can be used in.
They may also
need to carry information about inherent features that may matter to other phrases.
Section \ref{sec:writegf} contained an easy example with nouns. We used the categories \verb-N-
and \verb-N'- to separated between nouns with repsectively without a set number.
The full analysis includes a separation between the categories \verb-N- -- simple nouns --
and 
\verb-CN-, common noun phrases which may be modified by adjectival phrases or conjucted.
The \verb-N- category is implemented as an inflectiontable and field keeping information
about the gender.
A \verb-CN- may be turned into a \verb-NP- by the rule \verb-DetNP- which sets
the number and definitness. Since
all morphological information is generated by the smart paradigms and contained in the \verb-N-
the construction of a \verb-NP- more or less consists of putting the pieces together.
The two basic articles are \verb-IndefArt- and \verb-DefArt- which
uses the post-modified form of the noun: katten.
However, we need to consider the rules described in section \ref{sec:swedishnoun}. If we have
a common noun phrase consisting of the parts \emph{`liten'} (\emph{`small'}) and \emph{`katt'}
(\emph{`cat'}), there are three ways they can be combined:
\begin{verbatim}
DetNP : Det -> CN -> NP ;
        en    liten katt     -- DIndef
        den   lilla katten   -- DDef Def
        min   lilla katt     -- DDef Indef
\end{verbatim}        
>> compare NP to HPSG.
>> CN may be different in other theories.
>> how we have basic Pred, compl
>> clase is a pre-sentence needs tense, polarity etc.
Which combination that shoud be used is decided by the determiner. Hence all determiners
in our grammar must keep information of which definitness they allow. They record field
has a field storing this and the parameters are called \verb-DIndef-, 
\verb-DDef Indef- and \verb-DDef Def-. The determiners have arguments telling
which number the noun phrase should be in.
We also need to make sure that we do not allow constructions as given in
\ref{sent:trottkatt}a.
The definite article  \emph{``en"}is considered to be a quantifier, in plural it
is the either \emph{``de"} or nothing:
\begin{verbatim}
DetNP : Det              -> CN        -> NP ;
        definite, plural +  katt      =  katter
        definite, plural +  stor katt =  de stora katterna
\end{verbatim}        
This difference depends on the common noun phrase, which must carry information
about whether it has been modified or not.
The resource grammar have differenciate between quantifiers -- \verb-Quant-,
determiners \verb-Det- and predeterminers. Predeterminers modify \verb-NP-s
while the other to modifies \verb-CN-s. The differences are shown in table
\ref{tab:detquant}
\begin{figure}[h]
\begin{tabular}{l|lll}
                        &\textbf{Has number}& \textbf{Has definitness}& Example  \\ 
\hline
\textbf{Predeterminers} & --  & --  & alla katter, all maten \\ 
\textbf{Quantifiers}&-- &\checkmark  & min katt, mina katter, *min katten \\ 
\textbf{Determiners}   &\checkmark &\checkmark & varje katt, *varje katten  \\ 
\end{tabular}
\caption{}\label{tab:detquant}
\end{figure}

When a category has been formed, there is no longer any information available
about how it was put together. This is a result of the functional approach
of GF. In the case of the of the \verb-CN- above,
we would like to know whether
the it has been modified or not, that is, whether the rule
\verb|AdjCN : AP -> CN -> CN ;| has been applied. Insted the information
is passed on by an inherent feature, a flag telling set to tell
if the \verb-AdjCN- was applied.

In gf, only personal pronouns are considered to belong to PN. The rests, eg.
sådan (\emph{``such"}) are either Quantifiers or Dets.

In figure \ref{fig:gftree1} we saw the GF representation of the sentence ``Jag
ser katten". The verb ``såg" (see) takes one object: ``katten". Valency information
is encoded into the GF categories. The lexicon is therefore a very important part of the
grammar. 
\begin{verbatim}
see_V2 ; 
SlashV2a : V2 -> VPSlash ;
ComplSlash : VPSlash -> NP -> VP ;
\end{verbatim}
The simplest idea would let the VP consist of one field where ``se" to ``katten" is
glued together. But we also want to be able to express sentencens like
``Du ser inte mig" and ``Mig ser du inte", where the object is separated from the verb.
The \verb-VP- category is hence implemented to ressembel the satsschema (Diedriches,xx), 
it has multiple fields that may be put in the correct order later. There are fields
for negation, different types of adverbs, objects, the finite verb and an infinite verb.
The fields may in turn be tables showing the differnt forms they may have.

\begin{verbatim}
han (har   inte alltid tänkt på henne så)
     finit  neg  adV    inf  comp obj adv
\end{verbatim}

The funciton \verb-SlashV2a- lets you create a \textit{VPSlash}, a verb phrase missing a
noun phrase \cite{gazdar}. We combine the SlashVP with the object and get a complete
verb phrase.
\begin{verbatim}
 presens negative  (du ser mig)
  np.s ++ vp.fin ++ vp.neg ++ vp.obj ; 
\end{verbatim}

The slash-categories cat be combined with a subject to form a \verb-ClSlash-,
a clause missing a noun phrase. This is a convenient way to form questions ,
relative clauses
and topicalized clauses.
\begin{verbatim}
wh-questions: (interrogativ pronoun + ClSlash)
               vad                 (subject: Johan, vpslash: tittar på)
relative clauses: object + relative pronoun + ClSlash
                  katten     som             (subject: Johan, vpslash: tittar på)
topicalized clauses: object + ClSlash
                     katten  (subject: Johan, vpslash: tittar på)
\end{verbatim}

%Categories as V,V2,VS...
%Maybe talk about ställa.
Swedish verbs may take up to four arguments \cite[p. 53]{stymne}.
\enumsentence{Jag tar med mig den till honom}\label{sent:tamed}
These may be particles, reflexive object, indirect objects and direct objects.
In the grammmar these will be encoded in different ways. The particles are
given in the lexicon, as well as the prepositions that are choosen by the verbs.
The entry for sentence \ref{sent:tamed} looks like this, assuming that we already
have an constant \verb-take_V- which produces the correct conjugation for ``ta":
\verb|ta\_med\_V3 = dirV3 (reflV (partV (take\_V "med")) (mkPrep "till") ;|
The category \verb-V3- tells us that the verb takes two arguments (``den" and ``honom").
The preposition is set to ``till", reflV shows that the verb always is used with a reflexive
pronoun and partV gives the particle.
The fact that the choosen prepositions is given to the verb in the lexicon makes the 
parse tree visualisation algorithm put them together. See parse tree \ref{fig:tittapa}.
This is also the case for particles, see parse tree \ref{tittapart}, but even if
the visualizations look the same the two cases are treated differently inside the grammar
to keep the actual difference.
\enumsentence{På pojken tittar du \\ **På när du springer tittar jag}
As already stated, the visualized parse trees is not a complete represtation, and the reader
should not be confused or angry at this.\\

@@

In some special cases, phrase cannot be put in all forms. 
Give a good example!!
It is not quite clear how to handle this at the moment. When
the grammar is meant for parsing only, it is feasible to put
a \textit{nonexist} pattern.
example with this.\\

>> change dot file to have VP NP in correct order
\enumsentence{\shortexnt{8}
{Har& han& inte& ätit& de & gula& äpplena& idag?}
{Has& he& not& eaten& the& yellow& apples& today?}} 
\label{gfSwe:parseable}
The resource grammar can construct and parse the sentence in example
\ref{gfSwe:parseable}
and the parse tree is represented as in figure \ref{gfSwe:parsetree}.
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=70mm,height=70mm]{apples.pdf}
\caption{Parse tree for ``Har han inte ätit de gula äpplena idag".}
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit.."} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase. At the code level, this is implemented
using records.
The linearization rule for questions -- clauses with inverted order -- picks 
the field for the finite verb and appends it to the subject followed by
the negation, infinite part and the complement of the verb phrase: \\
\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\

Relative clauses are also covered: 
\enumsentence{\shortexnt{5}{Hon & ser & pojken & som  &sover}
              {she & sees & the boy & who & sleeps}}
              %{`She sees the sleeping boy'}}

\enumsentence{\shortexnt{6}{Han & ser &katten & han  &tycker & om}
              {he & sees & the cat & he & likes &}}
              %{`She sees the sleeping boy'}}
In addition to the basic resource grammar, additional constructions had been added
>> core is shared but particular lang. has particular features.
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
Among those were functions for topicalisation, although not all
of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
For example, \verb|FocObj| fronted the object as in sentence \ref{gfSwe:apple}.
\enumsentence{\shortex{6}
{Det & äpplet & vill & jag & inte & ha}
{that & apple & want & I & not &}
{`I don't want that apple'}}\label{gfSwe:apple}
The module provided functions for preposition stranding:
\enumsentence{\begin{tabular}{ll}
Stranded preposition & cf.\\
Vem måste jag akta mig för? & För vem måste jag akta mig?\\
Who do I need to watch out for? & For whom do I need to watch out?\\
\end{tabular}}

@@

\section{Development of the grammar}
\label{sec:Added}
The resource grammar is meant to cover the most important grammatical structures
present in many languages, but not the language specific constructions.
It has earlier been hard to identify missing constructions of the Swedish
implementation, since there was no big resource to evaluate it on.
When first conducting tests in this project, we found much room for improvement.
From the topcis listed in section \ref{sec:swedish}, the post-nominal artciles
and the verb-second property
were covered by the resources, as well as the periphrastic passive.
A start on topicalisation was implemented, while
the others are examples of what has been added during this project.
%enough to be covered by the resources. The others have been added.
%We have increased this by adding language specific
%constructions. Talbanken has been our inspiration.
Even though the grammar will not be totally covering, we belive that
the work done is both interesting and useful.
%>> the grammar will not be totally covering, but interesting and useful.
%>> hard to estimate before, since no big use case/resource.
%>> after initial experiments, first evaluation. found much room for improvement. 
%>> discovered constructions missing.

%%PASSIV
\subsection{The s-passive}
Passive voice is often used in Swedish, especially the
\textit{s-passive}.
\enumsentence{
\shortex{5}{Uppsatsen & skrev\textbf{s} & av & en & student.}
        {the essay & wrote+\textbf{s} & by & a & student}
        {`The essay was written by a student.'}}\label{sent:skrevs}
%{b. &En & student & skrev & uppsatsen.}
%        {&a &student &wrote &the essay}
%        {~~`A student wrote the essay.'}}
Some studies suggest that the s-passive is used in more than 80\% of the times \cite{laanemets}.
It is not as common in the other Scandinavian languages however, 
and it can not be used in all tenses for all words. The Norwegian 
translation of sentence \ref{sent:skrevs} is:
%The periphrastic passive is preferred, especially in spoken langauge.
\enumsentence{\shortexnt{7} %\begin{tabular}{ll}
{Oppgaven &ble& skrevet& av& en& student & [NO]}
{uppsatsen& blev& skriven& av& en & student& [SE]}}
%\end{tabular}}
The corresponding Swedish sentence is acceptable, but not as natural sounding as sentence
\ref{sent:skrevs}.
The resource grammar for Scandinavian therefore ~~expressed~~ the passive using auxiliary verb. \\
\verb|PassV2 : V2  -> VP | \\
\verb|         ta -> blev tagen| \\
%>> how are they used with reflexives??
This rules allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into complete verb phrases; they no longer need an object.
The s-passive has now been added as the standard case for the Swedish GF
grammar. Periphrastic passive is still allowed, but as an alternative rather
than the default.
The grammar also allows not only V2 but all verb phrases missing an object to 
form passives:
\begin{verbatim}
PassVPSlash : VPSlash -> VP ;
              ta      -> togs
              erbjöd  -> erbjöds
\end{verbatim}
\enumsentence{First place in two-place verbs\\\shortex{3}
{Jobbet & erbjöds & henne }
{\emph{the job} & \emph{offered+\textbf{s}}&\emph{ henne}}
{``The job was offered to her"}}
\label{ex:passV32}
\enumsentence{Second place in two-place verbs\\\shortex{3}
{Hon & erbjöds & jobbet}
{\emph{she} &\emph{ offered+\textbf{s}}&\emph{ the job}}
{``She was offered the job"}}
\label{ex:passV33}
\enumsentence{\shortex{3}
{Huset & målades & rött}
{\emph{the house} & \emph{painted+\textbf{s}}& \emph{red}}
{``The house was painted red"}}
\label{ex:passV2A}

@@

Transitive verbs may not be used like this
\enumsentence{\shortexnt{7}
{*Det & äter & en & fågel & frön & på & taket}
{~\emph{it} & \emph{eats} & \emph{a} & \emph{bird} & \emph{seeds} & \emph{on }& \emph{the roof}}}

>>>>

\chapter{The grammar}
\label{sec:prog.grammar}
An important part of this project has been to develope the Swedish GF grammar and
to adapt it to cover
constructions used in Talbanken. As a grammar implementation never can be
expected to give
full coverage of a language, we aim for a grammar fragment which gives a deep
analysis of the most important Swedish constructions.
The starting point has been the GF resource grammar and the new implementation is
still compatible with this.
Before describing the work progress we will give an introduction to the
resource grammars in general and to the Swedish implementation in particular.
This is found in section \ref{sec:swegf} while section \ref{sec:Added} shows and
explains the development during this project.

@@

\section{The Swedish resource grammar}
%>> start by easy examples, jag ser inte dig, du ser inte mig osv. 
%>> plocka från svenska delen 1 2 4 
%>> show simple trees
%>> explain all in Sw. section.
%>> show abstract and concrete tree, compare. (or in Gf sect)
%>> no subjects or object, analyse is more from the lexicon,
%>> compare to others, ask elisabet.\\
\label{sec:swegf}
%>> gf has pos-driven, leads to part-of-sentence analyse. first have v2,a2,vs etc.
%>> they may be combined in traditional way, and so the types depend and relate to each other
%>> see slides from summer school.
>> functions for predication, complementation.
>> rather different from others syntactical theories. does not remember syntactical
>> parts of sentences, like subj, obj, attributes.
%>> worked well until now, but particular langs. needs particular solutions. --> reflexives!
%>> no syntactic theory can cover a whole language, and so gf can't either.
>> shares 80\% with scandinavian, mostly words that differ.\\
>> complete morphology, limited syntax\\
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology,
word order, agreement, tense, basic conjunction etc.
 %and many phrase and clause constructions.
Due to the syntactic similarities of the Scandinavian languages, much of the implementation
is shared with Norwegian and Danish. The modules that concern the lexical
aspects are separate, while 85\% of the syntax description is shared.
There are about 80 functions, which describe the rules for building phrases and clauses.
\begin{verbatim}
PredVP       : NP -> VP -> Cl ; -- Prediciation
ComplVPSlash : VPSlash  -> VP ; -- Complementation
\end{verbatim}
The analysis preformed by GF is driven by the parts of speech, which are combined into
parts of sentences.
Figure \ref{fig:gfpic} shows the different categories, or types, used in the
resource grammars. 
A lexicon normally consists of words from the open word-classes, shown in rectangular boxes in
the picture. Each lexical entry is assigned with a type describing its word-class,
and the abstract grammar can be viewed as a set rules showing
%how the parts of speech can be used --
how the types relate to each other.
\begin{figure}[h]
\includegraphics[height=90mm]{categories.png}
\caption{The types of the GF resource grammars.}
\label{fig:gfpic}
\end{figure}

The concrete grammar gives a \textit{linearization type} for each category,
usually a record containing a table with the forms the phrase can be used in.
They may also
carry information about inherent features that matter to other phrases.

\subsection{Noun phrases}
Section \ref{sec:writegf} contained an easy example with nouns. We used the categories \verb-N-
and \verb-N'- to separated between nouns with repsectively without a set number.
The full analysis includes a separation between the categories \verb-N- -- simple nouns --
and 
\verb-CN-, common noun phrases which may be modified by adjectival phrases or conjucted.
The \verb-N- category is implemented as an inflection table, generated by the smart paradigm,
and field keeping information about the gender.
\begin{verbatim}
{s : Number => Species => Case => Str ; g : Gender}
flicka_N = {s = {Sg Indef Nom =>  flicka
                 Sg Indef Gen =>  flickas
                 Sg Def Nom   => flickan
                 Sg Def Gen   => flickans
                 Pl Indef Nom => flickor
                 Pl Indef Gen => flickors
                 Pl Def Nom   => flickorna
                 Pl Def Gen   => flickornas ;
            g = utr }
\end{verbatim}
When a noun is turned into a \verb-CN-, this information is kept, and the noun may
be modified.
\begin{verbatim}
AdjCN : AP -> CN -> CN ; 
\end{verbatim}
The function \verb-DetCN-, determination, creates a
\verb-NP- by setting the number and definitness of a \verb-CN-.
%may be turned into a \verb-NP- by the rule \verb-DetNP- which sets
%the number and definitness.
%Since all morphological information is generated by the smart paradigms and contained in the \verb-N-
%the construction of a \verb-NP- more or less consists of putting the pieces together.
%The definitness is given by a determiner.

\begin{figure}[h]
\begin{verbatim}
DetCN : Det                    -> CN     -> NP ;
        indefinitive, singular -> flicka -> "en flicka"
        definitive,singular    -> flicka -> "flickan"
\end{verbatim}        
\caption{}
\label{gfcode:DetCN}
\end{figure}

%The two basic articles are \verb-IndefArt- and \verb-DefArt- which
%uses the post-modified form of the noun: katten.
Some rules for determination were described in section \ref{sec:swedishnoun};
if we have a common noun phrase consisting of the parts \emph{`liten'}
(\emph{`small'}) and \emph{`katt'} (\emph{`cat'}), there are three ways they
can be combined:
\begin{figure}
\begin{tabular}{|lll|l|}
\hline
\textbf{Determiner} & \textbf{Adjective} & \textbf{Noun} & \textbf{DetSpecies}\\
\hline
en  &  liten $[${\sc -Def}$]$& katt   $[${\sc -Det}$]$& \textbf{ DIndef}\\
min &  lilla $[${\sc +Def}$]$& katt   $[${\sc -Det}$]$  & \textbf{ DDef Indef} \\
den &  lilla $[${\sc +Def}$]$& katten $[${\sc +Det}$]$& \textbf{ DDef Def} \\
\hline
\end{tabular}        
\caption{}\label{tab:detdef}
\end{figure}

%Which combination that shoud be used is decided by the determiner. Hence all
Hence, all determiners
in our grammar must keep information of which definitness they allow, the
\verb-DetSpecies- parameter. The number is also stored as an inherent 
feature of the determiner.
The resource grammar differenciates between quantifiers, 
determiners \verb-Det- and predeterminers. Predeterminers modify \verb-NP-s
while the other to modifies \verb-CN-s. The differences are shown in table
\ref{tab:detquant}. 
\begin{figure}[h]
\begin{tabular}{l|lll}
                        &\textbf{Has number}& \textbf{Has definitness}& Example  \\ 
\hline
\textbf{Predeterminers} & --  & --  & alla katter, all maten \\ 
\textbf{Quantifiers}&-- &\checkmark  & min katt, mina katter, *min katten \\ 
\textbf{Determiners}   &\checkmark &\checkmark & varje katt, *varje katten  \\ 
\end{tabular}
\caption{}\label{tab:detquant}
\end{figure}

The definite article is considered to be a quantifier, which has the forms
\emph{en} and \emph{ett} for singular in plural it
is the either \emph{``de"} or nothing, depending on whether the noun phrase
has been modified.
This information must thus be stored in the \verb-NP-. 
\begin{figure}[h]
\begin{verbatim}
DetNP : Det              -> CN        -> NP ;
        definite, plural +  katt      =  katter
        definite, plural +  stor katt =  de stora katterna
\end{verbatim}        
\caption{}\label{gfcode:defmod}
\end{figure}

%Their record
%has a field storing the
%which may  are called \verb-DIndef-, \verb-DDef Indef- and \verb-DDef Def-. 
%Additionally, we need to make sure that sentences as 
%\ref{sent:trottkatt}a are not allowed.

Many words which are considered to be pronouns in other analyses are 
classified as quantifiers and determiners in GF.
Only personal pronouns are of the type \verb-Pron-.

>> compare NP to HPSG.
>> CN may be different in other theories.
---------
When a category has been formed, there is no longer any information available
about how it was put together. This is a result of the functional approach
of GF. In the case shown in figure \ref{gfcode:defmod},
we would like to know whether
the it has been modified or not.
%that is, whether the function \verb|AdjCN| has been applied.
This information
has to be passed on by an inherent feature, a flag set to tell
if the \verb-AdjCN- was applied.

\subsection{Verb phrases}
In figure \ref{fig:gftree1} we saw the GF representation of the sentence ``Jag
ser katten". The verb ``såg" (see) takes one object: ``katten".
In GF, this can be seen on the type of the verb,
valency information
is encoded into the lexicon. Transitive verbs have the type \verb-V2-, ditransitive
\verb-V3- etc. There are types for verb which takes sentences (\verb-VS-) and
adjectival (\verb-VA-) complements.
The lexicon is therefore a very important part of the
grammar. 
\begin{verbatim}
see_V2 ; 
SlashV2a   : V2 -> VPSlash ;
ComplSlash : VPSlash -> NP -> VP ;
\end{verbatim}

The funciton \verb-SlashV2a- lets you create a \verb-VPSlash-, a verb phrase missing a
noun phrase \cite{gazdar}. We combine the \verb-SlashVP- with the object and
get a complete verb phrase.
Both the verb and its complements are stored in the verb phrase category, \verb-VP-.
%A simple idea would let the \verb-VP- consist of one field where the verb and object are
%glued together. But we also want to be able to express sentencens like
%``Du ser inte mig" and ``Mig ser du inte", where the object is separated from the verb.
The \verb-VP- category resembles the Didrichsen's field model,
it has multiple fields that may be put in the different orders. There are fields
for negation, different types of adverbs, objects, the finite verb and an infinite verb.
The fields may in turn be tables showing the differnt forms of the component.\\

\begin{tabular}{|l|lllllll|}
\hline
\textbf{Subj} && &\textbf{VP}& & & & \\
    & \sc finit& \sc neg& \sc adV & \sc inf& \sc comp& \sc obj& \sc adv \\
\hline
han & har &  inte&  alltid&  tänkt&  på&  henne&  så \\
\hline
\end{tabular}\\

\vspace{5mm}
The fields form the verb phrase are not put in the correct order until a sentence is
created.\\

Swedish verbs may take up to five arguments \cite[p. 53]{stymne}.
\enumsentence{Jag tar med mig den till honom}\label{sent:tamed}
These may be prepositions, particles, reflexive object, indirect objects and direct objects.
In the grammmar these will be encoded in different ways. The particles are
given in the lexicon, as well as the prepositions that are choosen by the verbs.
The entry for `\emph{ta med}' as used in sentence \ref{sent:tamed}, is described as follows:
%, assuming that we already have an constant \verb-take_V- which produces the correct conjugation for ``ta":
\begin{verbatim}
ta_med_V3 = dirV3 (reflV (partV (take_V "med")) (mkPrep "till") ;
\end{verbatim}
The category \verb-V3- tells us that the verb takes two objects, % (``den" and ``honom").
the second one is to be used with the preposition `\emph{till}'. \verb-reflV- shows
that the verb always is used with a reflexive
pronoun and \verb-partV- gives the particle.
The fact that the choosen prepositions is attatched to the verb in the lexicon
causes the parse tree visualisation algorithm group them together.
This is also the case for particles, cf. parse tree \ref{pic:tittapart}
and \ref{pic:tittapa}.
As already stated, the visualized parse trees is not a complete represtation,
even if the verb phrases in the the visualizations look the same, the two cases
are treated and represented differently internally. The fronting of the
preposition as in of sentence \ref{sent:tittapa}a. is accepted but
not fronting of particles as in\ref{sent:tittapa}b.
\eenumsentence{
\item[a]På pojken tittar du 
\item[b] *På när du springer tittar jag}\label{sent:tittapa}.
\begin{figure}[h]
\subfloat[Verb with a chosen preposition]{\label{pic:tittapa}\includegraphics[height=80mm]{tittapa.pdf}}
\hspace{10mm}
\subfloat[Verb with particle]{\label{pic:tittapart}\includegraphics[height=90mm]{tittapart.pdf}}
\caption{The visualized parse trees do not show the internal difference of chosen prepositions and particles}
\label{fig:translationtrees}
\end{figure}

%Categories as V,V2,VS...
%Maybe talk about ställa.
\subsection{Clauses}
The category clause, \verb-Cl-, represents a pre-sentence, that does not yet have any tense, polarity
or word order set.\\

\begin{tabular}{|lll|ll|}
\hline
\textbf{Tense} &\textbf{Polarity} &\textbf{Word order} &\textbf{Cl} &\textbf{S} \\
\hline
Presens &Negative & Main        & du ser mig & 'Du ser inte mig'\\
Perfect &Positive & Inverted    & du ser mig & 'Har du sett mig'\\
Perfect &Negative & Subordinate & du ser mig & 'Du har sett mig'\\
\hline
\end{tabular}\\

Like verb phrases, a clause may also be missing an object and then has the type
\verb-ClSlash-. The \verb-ClSlash- is formed by a \verb-VPSlash- which is 
given a subject.
This is a convenient way to form questions , relative clauses
and topicalized clauses.\\

\begin{tabular}{|l|ll|}
\hline
\textbf{Cl}& Johan + tittar på \\
\hline
\textbf{Wh-Questions} & Interrogativ pronoun   & Vad tittar Johan på?\\
\textbf{Relative clauses} & Object + Relative pronoun & Katten som Johan tittar på \\
\textbf{Topicalized clauses} & Object  & Henne tittar Johan på \\
\hline
\end{tabular}\\

@@

%In some special cases, phrase cannot be put in all forms. 
%Give a good example!!
%It is not quite clear how to handle this at the moment. When
%the grammar is meant for parsing only, it is feasible to put
%a \textit{nonexist} pattern.
%example with this.\\

\subsection{Overview}
\vspace{4mm}
The resource grammar could express complex sentences
as the one in figure \ref{gfSwe:parsetree}.
%shows the parse tree for sentence 
%\ref{gfSwe:parseable}.\\
>> change dot file to have VP NP in correct order
%\enumsentence{\shortexnt{8}
%{Har& han& inte& ätit& de & gula& äpplena& idag?}
%{Has& he& not& eaten& the& yellow& apples& today?}} 
%\label{gfSwe:parseable}
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=70mm,height=70mm]{apples.pdf}
\caption{Parse tree for ``Har han inte ätit de gula äpplena idag".}
\label{gfSwe:parsetree}
\end{figure}\\
Even though the verb phrase \emph{``har inte ätit.."} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase.
%At the code level, this is implemented %using records.
%The linearization rule for questions -- clauses with inverted order -- picks 
%the field for the finite verb and appends it to the subject followed by
%the negation, infinite part and the complement of the verb phrase: \\
%\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\

The resource grammar covered relative clauses: 
\eenumsentence{\item[a]\shortexnt{5}{Hon & ser & pojken & som  &sover}
              {\emph{she} & \emph{sees} & \emph{the boy} & \emph{who} & \emph{sleeps}}
              %{`She sees the sleeping boy'}}
\item[b]\shortexnt{6}{Han & ser &katten & han  &tycker & om}
              {\emph{he} & \emph{sees} & \emph{the cat} & \emph{he} & \emph{likes} &}}
              %{`She sees the sleeping boy'}}

In addition to the core of the resource grammars, which is shared with the other languages implemented in the
library, 
%In addition to the basic resource grammar,
 language specific constructions may bee added
%>> core is shared but particular lang. has particular features.
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
Among those were functions for topicalisation: %$ although not all
%of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
%For example, \verb|FocObj| fronted the object as in sentence \ref{gfSwe:apple}.
\enumsentence{\shortex{6}
{Det & äpplet & vill & jag & inte & ha}
{that & apple & want & I & not &}
{`I don't want that apple'}}\label{gfSwe:apple}
and for preposition stranding:
\enumsentence{\begin{tabular}{ll}
Stranded preposition & cf.\\
Vem måste jag akta mig för? & För vem måste jag akta mig?\\
Who do I need to watch out for? & For whom do I need to watch out?\\
\end{tabular}}

@@

\section{Development of the grammar}
\label{sec:Added}
%The resource grammar is meant to cover the most important grammatical structures
%present in many languages, but not the language specific constructions.
It has earlier been hard to identify missing constructions of the Swedish
implementation, since there was no large resource available to evaluate it on.
When first conducting tests in this project, we found much room for improvement.
From the topcis listed in section \ref{sec:swedish}, the post-nominal artciles
and the verb-second property
were covered by the resources, as well as the periphrastic passive.
A start on topicalisation was implemented, while
the others are examples of what has been added during this project.
%enough to be covered by the resources. The others have been added.
%We have increased this by adding language specific
%constructions. Talbanken has been our inspiration.
%>> the grammar will not be totally covering, but interesting and useful.
%>> hard to estimate before, since no big use case/resource.
%>> after initial experiments, first evaluation. found much room for improvement. 
%>> discovered constructions missing.

%%PASSIV
\subsection{The s-passive}
Passive voice is often used in Swedish, especially the
\textit{s-passive}.
\enumsentence{
\shortex{5}{Uppsatsen & skrev\textbf{s} & av & en & student.}
        {the essay & wrote+\textbf{s} & by & a & student}
        {`The essay was written by a student.'}}\label{sent:skrevs}
%{b. &En & student & skrev & uppsatsen.}
%        {&a &student &wrote &the essay}
%        {~~`A student wrote the essay.'}}
Some studies suggest that the s-passive is used in more than 80\% of the times
\cite{laanemets}.
It is not as common in the other Scandinavian languages however, 
where not all words have passive forms for all tenses. The Norwegian 
translation of sentence \ref{sent:skrevs} is:
%The periphrastic passive is preferred, especially in spoken langauge.
\enumsentence{\shortexnt{7} %\begin{tabular}{ll}
{Oppgaven &ble& skrevet& av& en& student & [NO]}
{uppsatsen& blev& skriven& av& en & student& [SE]}}
%\end{tabular}}
The corresponding Swedish sentence is acceptable, but not as natural sounding as sentence
\ref{sent:skrevs}.
The resource grammar for Scandinavian therefore implemented the function for passive,
\verb-PassV2-, by using auxiliary verb. \\
\begin{verbatim}
PassV2 : V2  -> VP ;
         ta -> blev tagen
\end{verbatim}
%>> how are they used with reflexives??
The function allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into complete verb phrases; they no longer need an object.\\

During this project, the s-passive was added as the standard case for the Swedish GF
grammar. Periphrastic passive is still allowed, but as an alternative rather
than the default.
The grammar further allows not only V2 but all verb phrases that misses an object to 
form passives:
\begin{verbatim}
PassVPSlash : VPSlash -> VP ;
              ta      -> togs
              erbjöd  -> erbjöds
\end{verbatim}
\enumsentence{\textbf{First place in two-place verbs}\\\shortex{3}
{Jobbet & erbjöds & henne }
{\emph{the job} & \emph{offered+\textbf{s}}&\emph{ henne}}
{``The job was offered to her"}}
\label{ex:passV32}
\enumsentence{\textbf{Second place in two-place verbs}\\\shortex{3}
{Hon & erbjöds & jobbet}
{\emph{she} &\emph{ offered+\textbf{s}}&\emph{ the job}}
{``She was offered the job"}}
\label{ex:passV33}
\enumsentence{\textbf{V2A verb}\\\shortex{3}
{Huset & målades & rött}
{\emph{the house} & \emph{painted+\textbf{s}}& \emph{red}}
{``The house was painted red"}}
\label{ex:passV2A}

@@

Transitive verbs may not be used like this
\enumsentence{\shortexnt{7}
{*Det & äter & en & fågel & frön & på & taket}
{~\emph{it} & \emph{eats} & \emph{a} & \emph{bird} & \emph{seeds} & \emph{on }& \emph{the roof}}}
\label{sent:fagelfron}
